---
title: "Marine Mammal Stock Assessment Data"
subtitle: "Public, reproducible SAR data — built from GAMMS IV"
date: today
toc: false
resources:
  - data/stock_summary.csv
  - data/data_dictionary.csv
  - data/fisheries_msi.csv
---

```{r}
#| label: setup
#| include: false
library(readr)
library(dplyr)
library(reactable)
library(htmltools)

stocks <- read_csv("data/stock_summary.csv", show_col_types = FALSE)

# Reproducibility: derive n_min and PBR from the raw inputs using the GAMMS
# formulas, rather than trusting stored values. The portal shows the COMPUTED
# numbers, so the data can never silently disagree with the guidelines.
stocks <- stocks |>
  mutate(
    n_min_calc = n_best / exp(0.842 * sqrt(log(1 + cv_n^2))),
    pbr_calc   = n_min_calc * 0.5 * r_max * f_r
  )
```

```{r}
#| label: lastmod
#| include: false
n_stocks <- nrow(stocks)
```

This portal turns marine mammal Stock Assessment Reports (SARs) into structured,
downloadable data. Every value follows the
[GAMMS IV guidelines](https://www.fisheries.noaa.gov/national/laws-and-policies/policy-directive-system).
**`r n_stocks` stocks** are currently available. *(Data shown are illustrative
placeholders during development.)*

```{r}
#| label: table
#| echo: false
reactable(
  stocks |>
    transmute(
      Species   = sprintf("%s (%s)", common_name, scientific_name),
      Stock     = stock_name,
      Region    = region,
      Year      = sar_year,
      Abundance = round(n_best),
      CV        = round(cv_n, 2),
      Nmin      = round(n_min_calc),
      PBR       = round(pbr_calc, 1),
      `Annual M/SI` = annual_msi_total,
      Status    = strategic_status,
      ESA       = esa_status
    ),
  searchable = TRUE,
  filterable = TRUE,
  sortable   = TRUE,
  defaultPageSize = 15,
  highlight  = TRUE,
  striped    = TRUE,
  defaultColDef = colDef(align = "left", minWidth = 70),
  columns = list(
    Species   = colDef(minWidth = 160),
    Abundance = colDef(format = colFormat(separators = TRUE), align = "right"),
    Nmin      = colDef(format = colFormat(separators = TRUE), align = "right"),
    PBR       = colDef(align = "right"),
    `Annual M/SI` = colDef(align = "right"),
    Status = colDef(cell = function(value) {
      color <- if (value == "strategic") "#A32D2D" else "#3B6D11"
      bg    <- if (value == "strategic") "#FCEBEB" else "#EAF3DE"
      span(style = sprintf("color:%s;background:%s;padding:2px 8px;border-radius:6px;font-size:12px", color, bg), value)
    })
  )
)
```

## How to access this data

::: {.grid}

::: {.g-col-12 .g-col-md-4}
### Download
- [stock_summary.csv](data/stock_summary.csv)
- [fisheries_msi.csv](data/fisheries_msi.csv)
- [data_dictionary.csv](data/data_dictionary.csv)
:::

::: {.g-col-12 .g-col-md-4}
### Read directly in R
```r
url <- paste0(
  "https://raw.githubusercontent.com/",
  "isabellay02/PSAWtraining/main/",
  "data/stock_summary.csv")
stocks <- read.csv(url)
```
:::

::: {.g-col-12 .g-col-md-4}
### What the fields mean
Every column — units, definition, and the GAMMS section it comes from — is
documented in the [data dictionary](data/data_dictionary.csv).
:::

:::
